---
title: "Simulation for the Effects of Social Media on Mental Health"
author: "Haley Lam"
date: "`r Sys.Date()`"
output: 
  html_document: 
    theme: cosmo
---

### Overview

This is a simulation of what happens to outputs with typical OLS regression on cross-sectional data versus a quasi-experimental design (Difference-in-differences). The DiD design is heavily inspired by Braghieri et al. (2022), and the framework follows Baker et al. (2025). Monte carlo simulations follow Cunningham (2021)'s approach to teaching about causal inference.


## 1. Set Envrionment and Set Parameters

```{r}
library(ggplot2)
library(tidyverse)
```

```{r}
N          <- 200     # Total colleges
delta      <- 0.085   # True causal effect
gamma_u    <- 0.30    # Unobserved covariate - OLS has no data for this
gamma_w    <- 0.15    # Observed covariate - OLS will 'control' for this
time_trend <- 0.05    # Common time trend
```

## 2. Make Some Data!

The idea from Braghieri et al. (2022): Facebook was (randomly) rolled out to colleges at different times. We can use this exogenous shock to examine the effects of its availability on mental health outcomes. Therefore, our unit analysis here will be colleges (i.e., average college pre vs. post facebook)

```{r}
set.seed(123)

# Create the 200 colleges and their traits
colleges <- tibble(
  college = 1:N,
  Treat   = rep(c(1, 0), each = N / 2),
  U       = rnorm(N),
  W       = rnorm(N) 
)

df <- colleges %>%
  crossing(period = 1:2) %>% # Duplicate each college to have both time 1 and time 2
        mutate(
                Post = as.integer(period == 2),
                D    = Treat * Post,  # 1 only for treated colleges in the post-period
                # Generate random error terms
                e_sm = rnorm(n(), mean = 0, sd = 0.5),
                e_mh = rnorm(n(), mean = 0, sd = 0.5),
                # Generate Social Media (SM) use
                SM = gamma_w * W + gamma_u * U + D + e_sm,
                # Generate Mental Health (MH) outcomes (higher = worse)
                MH = delta * SM + gamma_u * U + gamma_w * W + (time_trend * Post) + e_mh
  )

head(df)
```

## 3. Examine the Differences in Differences by Four-Means Table

First, we will calculate the Difference-in-Differences "by hand" using a simple Four-Means table (Baker et al., 2025). Then, we will compare that result against standard OLS regressions to see why cross-sectional data fails us here.

```{r}
# 1. Four-Means Table (Baker et al., 2025)
# a simple table that shows means for each group and period
four_means <- df %>%
  mutate(
    Group  = ifelse(Treat == 1, "Treated", "Control"),
    Period = ifelse(Post == 1, "Post", "Pre")
  ) %>%
  group_by(Group, Period) %>%
  summarize(mean_MH = mean(MH), .groups = "drop") %>%
  pivot_wider(names_from = Period, values_from = mean_MH) %>%
  mutate(Change_Over_Time = Post - Pre)

print(four_means)

# 2. Calculate DiD Estimate From the Table
delta_T <- four_means$Change_Over_Time[four_means$Group == "Treated"]
delta_C <- four_means$Change_Over_Time[four_means$Group == "Control"]
did_manual <- delta_T - delta_C

round(did_manual, 4) # This is the estimated causal effect from the DiD, descriptively. 
```

## 4. Compare OLS vs. DiD Regressions
```{r}
# For standard OLS, we only look at the post-treatment cross-sectional data
df_post <- df %>% filter(Post == 1)

# Regression 1: Naive OLS (Just SM and MH)
ols_naive <- lm(MH ~ SM, data = df_post)

# Regression 2: Adjusted OLS (Controlling for the observed variable 'W', like anxiety)
ols_adj <- lm(MH ~ SM + W, data = df_post)

# Regression 3: DiD Regression (Using the full panel data and interaction term)
did_reg <- lm(MH ~ Treat * Post, data = df)

cat("1. OLS (Naive) SM Coefficient     :", round(coef(ols_naive)["SM"], 4), "\n")
cat("2. OLS (Adjusted) SM Coefficient  :", round(coef(ols_adj)["SM"], 4), "\n")
cat("3. DiD (Treat:Post) Coefficient   :", round(coef(did_reg)["Treat:Post"], 4), "\n")
cat("True Causal Effect (delta)        :", delta, "\n")
```

As you can see, the estimate from the DiD is much closer to the true causal effect.
