---
title: "Portfolio 4: A DAG Structure for Social Media's Impact on Mental Health"
author: "Haley Lam"
date: "`r Sys.Date()`"
output: 
  html_document: 
    theme: cosmo
---

## Overview
The goal of this portfolio is to unpack a causal structure as informed by the literature, intended for readers who want the complete picture. We'll start from a simple DAG to more complicated structures to show the value of thinking through a causal structure before running analyses. There will be three panels, each with a threat to causal inference:
Panel A: Two unmeasured confounders and observed covariates.
Panel B: Conditioning on a Collider
Panel C: Conditioning on a Mediator

Let's do it!

### Load Libraries (there's no data to load!)

```{r}
library(ggplot2)
library(patchwork)
library(dagitty)
```

## Section 1: Using "daggity" to draw and analyze DAGs

Daggity is a tool (and a package in R) that will help us understanding our hypothesized underlying causal structure. They have a website, but since we're in R, we can use their package!

### DAG #1: Unmeasured Confounding

These are the variables we'll be dealing with: 
SM  = Social media use (treatment)
MH  = Mental health outcome (depression/wellbeing)
U1  = Neuroticism / pre-existing mental health (unobserved latent)
      Predicts heavier SM use (emotional regulation, boredom avoidance)
      AND worse MH outcomes (pre-existing vulnerability). 
U2  = SES / family environment (unobserved latent)
      Lower SES predicts higher SM use (indoor leisure) AND worse MH (material hardship, neighborhood stress). 
      There are proxies in W but they're not fully captured.
W   = Observed covariates (age, sex, race, parental education)
      Standard controls in the correlational literature. Reduce bias from demographic variation but do NOT close the U1 or U2 backdoor paths.

#### Daggity for Dag #1

Unmeasured confounding is something psychologists are aware about, and possibly the main reason why psychologists don't trust in causality with observational data. Here, we show how it might appear in a DAG. We first define the variables (also called nodes), then draw the arrows (or edges).

```{r}
dag_A <- dagitty('dag {
  SM 
  MH 
  U1 [latent]
  U2 [latent]
  W
  
  SM -> MH
  U1 -> SM
  U1 -> MH
  U2 -> SM
  U2 -> MH
  W  -> SM
  W  -> MH
}')
```

What we did above was define our causal structure. There are many ways to do this, including forks (SM <- W -> MH) and chains (we don't have one here, but it'd look something like A -> B -> C), but what I did above was the simplest way, which defines causal directions with just pairs of variables. Now, we'll use the "adjustmentSets" function to "test" our DAG! The output tells us what we must control for.

We can first test a minimal solution. We should get an empty response, since we have unmeasured (latent) confounders that cannot be controlled for. 
```{r}
print(adjustmentSets(dag_A, "SM", "MH", type = "minimal")) 
```

This means that there is no solution to our issue of unmeasured confounders! We're doomed! This is extremely useful to know, since we can control for our observed confounders (W) all we want, but this output tells us that with the data we currently have, an unbiased regression is impossible.

Now, let's see what we have to control to get an unbiased regression (i.e., block all backdoor paths). We can do this with the "type = "all"" setting.
```{r}
print(adjustmentSets(dag_A, "SM", "MH", type = "all"))
```

The ouput returned all adjustment sets possible. As we can see, we must adjust for U1, U2, AND W. Adjusting for only one or two of those would not work. In this first case, we learned that given our DAG has potential unobserved confounders, we have a biased regression if we simply condition on the observed confounder (i.e., the "try out best to control all we have" solution does not work!)

### DAG #2: Collider Bias
Here are our variables:
SM  = Social media use
MH  = Mental health
U   = Unobserved background confounders (neuroticism, pre-existing MH)
C   = In-person social interaction (the hypothesized collider)

C is a collider because both X and Y affect it:
SM → C : Substitution — heavy social media use displaces time that would otherwise be spent face-to-face (more SM → less in-person)
MH → C : Social withdrawal — depression and anxiety reduce motivation and ability to engage in in-person social activities (worse MH → less in-person interaction)

Since C has two causes on the path of interest (SM and MH), C is a collider on the path SM → C ← MH. The standard DAG result: conditioning on a collider *opens* a path between its causes. So including C as a covariate does not "control for" in-person socializing in a neutral way, but introduces a spurious relationship between SM and MH.

#### Daggity for DAG #2

Again, let's first define our DAG.

```{r}
dag_B <- dagitty('dag {
  SM 
  MH 
  U  [latent]
  C  
  
  SM -> MH
  U  -> SM
  U  -> MH
  SM -> C
  MH -> C
}')
```

We can do the same thing here with seeing minimal adjustment sets for SM->MH. Meaning, what can we control for given the data we have? 
```{r}
print(adjustmentSets(dag_B, "SM", "MH", type = "minimal"))
```

Well, nothing! There's nothing we can do! What if we had all data? What would we control for?
```{r}
print(adjustmentSets(dag_B, "SM", "MH", type = "all"))
```

Notice here that "C" is not in the output! This means that there is no case in which controlling for the collider would produce an unbiased estimator. 


Now, we can more clearly see the issue with controlling for a collider here. We can compare the before and after of 1) not controlling for the collider and 2) controlling for the collider.
```{r}
paths(dag_B, "SM", "MH", Z = c("U"))

paths(dag_B, "SM", "MH", Z = c("U", "C"))
```
We can see the in the first equation, SM -> C <- MH, is blocked (or closed). This is a good thing. We correctly control for U, and our only open path is the direct effect of SM -> MH. This is perfect!

However, in the second one, we opened the path SM -> C <- MH. This introduces bias and we now have an opened backdoor path! We now have some additional spurrious association betweeen SM and MH. 



